---
title: "R Code Basics"
subtitle: "BDiB Session 11-February-2021"
author: "Axel Nohturfft"
date: "`r Sys.Date()`"
output: 
  html_notebook:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 3
    code_folding: "hide"
---

Above: "YAML header"

<style>
h1 {background: brown;color: white;padding-left: 7px;}
h2 {background: grey;color: white;padding-left: 7px;}
h3 {color: brown;}
code {color: chocolate;}
</style>


## Scripts  
***ALWAYS SAVE YOUR CODE IN SCRIPTS !***  

* To document your work (think of them as a data analyst's lab notebook)
* To quickly repeat an analysis in the future
* To re-use your code for new analyses
* To generate reports (rmarkdown scripts)

There are two types of R scripts:  

* Standard R scripts (see `basics.R`)  
* rmarkdown (`.Rmd`) scripts  
* Please see the `rmarkdown.Rmd` or `rmarkdown.nb.html` files for details on rmarkdown scripts  

## Data structures (1): scalars and vectors  

* Vectors are simple lists with each item being of the same type.
* Scalars are really just vectors with a single item.
* Some useful vectors are 'built' into R:
* (place the cursor anywhere in a line of code and press Ctrl-Enter to execute
* the code / Cmd-Enter on a Mac)
* Results / output will appear in the 'Console' window below the script window.



```{r}
letters
```

```{r}
LETTERS
```

```{r}
month.abb
```

```{r}
month.name
```


Vectors can be generated using the c() function:
```{r}
c(14, 6, 2016, 12)
```

## Functions  
***Functions are used to manipulate data*** 

### Standard arithmetic  
R includes all the standard arithmetic function symbols: + - * / ^  
```{r}
3 + 4
```

```{r}
300 / 100
```

```{r}
10 - 9
```

```{r}
2 ^ 4
```

Modulo:  
```{r}
12 %% 5
```

### Names functions  
Named R functions are very similar to worksheet functions in Excel:  

1. Each function has a different name  
2. Function names are always followed by brackets  
3. Information typed between the brackets are called "arguments"  
4. Two or more arguments are separated by commas  

To see a list of all (>1200) built-in functions that come with R (aka "base R") type:  
```{r}
help(package="base")
```

So `help()` is a first example of a function. ' package="base" ' is the argument here.

### Some examples of frequently used functions  
(note how in some the examples we use vectors as arguments)  

```{r}
c(3, 4, 10)
```

```{r}
sum(c(3, 4, 10))
```

```{r}
sqrt(121)
```

```{r}
mean(c(3, 4, 10))
median(c(3, 4, 10))
max(c(3, 4, 10))
min(c(3, 4, 10))
```
## Variables  
Usually data are stored in named VARIABLES.  

* Variables are defined with the '<-' assignment operator  
* Variable names can be very long  
* Variables CANNOT contain spaces  
* Variables CANNOT contain dashes/hyphens    
* Variables CANNOT start with a number (but numbers can appear elsewhere)  

# Two scalars:
```{r}
my.first.variable <- "Hello world!"
my.second.variable <- 2019
```

***Note that variables are listed in the 'Environment' tab in the top right RStudio panel.***  

Use the 'print()' or 'cat()' function to display the contents of a variable.  
Output from `print()` is preceded by row numbers in square brackets; with `cat()` the row numbers are not printed:
```{r}
print(my.first.variable)
cat(my.first.variable)
```

```{r}
print(my.second.variable)
cat(my.second.variable)
```

Vectors can be generated using the `c()` function:
```{r}
v1 <- c(14, 6, 2016)
```

Typing just the variable is equivalent to `print()`:
```{r}
v1
```

The `View()` function opens a new tab - mostly used to inspect large tables.  
The tab will include several useful ways of inspecting and filtering a table.
```{r}
View(mtcars)
```

## Playing with vectors  
**Note**: The objective section is just to illustrate the usefulness of functions and programming languages in general. You do ***not*** need to remember any of the specifics!  

### Playing with text vectors  
```{r}
v2 <- c("Hello", "world", "!")
v2
```

Vector items can be 'glued' together into a single piece of text using the `paste()` function:
```{r}
paste(v2, collapse=" ")
```

Modifying capitalization:  
```{r}
toupper(v2)
tolower(v2)
```

### Playing with number vectors  
It's easy to generate vectors with simple series of numbers:

```{r}
v3 <- 1:12
v3
```

The seq command can generate more complicated number vectors:
```{r}
seq(from=1, to=22, by=3)
```

```{r}
vv <- seq(from=1, to=10, length.out=19)
vv
```

The `length()` function returns the number of items in a vector:  
```{r}
length(vv)
```

### Plotting vectors  
Plotting vectors is very straightforward:  
```{r}
plot(1:50)
```

```{r}
t <- (1:100) / 10
plot(sqrt(t) * cos(t), sqrt(t) * sin(t), type="l", axes=FALSE, col="blue")
```


### Vector items can have names  
The `names()` function is used both to define and to return the names of a data object.
```{r}
v3
month.abb
```

```{r}
names(v3) <- toupper(month.abb)
v3
```

```{r}
names(v3)
```

In R it is very easy to manipulate all items in a vector with simple code:  
```{r}
v3 * 10
```

## Subsetting (1): Subsetting vectors  

### Subsetting vectors by index  

You can select items from within a vector using square brackets + indices ...  

```{r}
# LETTERS is a built-in variable that stored the uppercase letters of the alphabet:
LETTERS
```


Subsetting with a single index number...  
```{r}
LETTERS[11]
```

Subsetting with a vector of index numbers ...  
```{r}
# The c() function generates a vector:
c(8,5,12,12,15)
```

```{r}
h <- LETTERS[c(8,5,12,12,15)]
h
```

Remember how to 'glue' vector items together?
```{r}
paste(h, collapse="")
```

### Subsetting vectors using names  
You can also refer to vector items by name, if defined ...
```{r}
v3
```

```{r}
# v3[11]
v3["NOV"]
# v3["NOV"] * 10000
```

Using a vector of names:
```{r}
# c("JAN", "MAR")
# v3[c(1,3)]
v3[c("JAN", "MAR")]
```

### Subsetting vectors using Boolean (TRUE/FALSE) values  
```{r}
v4 <- 1:3
v4
```

```{r}
c(TRUE, FALSE, TRUE)
v4[c(TRUE, FALSE, TRUE)]
```

It can be easier to save a vector in a variable first, then use the variable for subsetting:
```{r}
v4
v4 > 1
x <- v4 > 1
x
v4[x]
```
## Data types in R  

As we have already seen the most important data TYPES in R are:  
1. numeric  
2. text (aka 'character' or 'strings')  
3. Boolean (aka 'logical')  

*(numeric data are split further into sub-categories but we don't to worry
about that here)*  

### Numerical data  
```{r}
v3
is.numeric(v3)
class(3.5)
# typeof(v3) # integer = whole numbers
# typeof(3.5) # "double"
```

### Text data  
Text format in R is called 'character':  
```{r}
v2
class(v2)
# is.character(v2)
# is.numeric(v2)
# typeof(v2)
```

### Boolean/logical data  
Boolean/logical values are often generated through 'equal' or 'greater/lesser than' (<==>) operations:  

```{r}
10 > 2
2 > 10
3 == 3.0
4 == 5
```
The following generates a vector of boolean values:
```{r}
1:10         # generates a vector of numbers from 1 to 10
1:10 >= 5    # generates a vector of boolean values
```

Store in a vector:  
```{r}
v5 <- 1:10 >= 5
v5
```

To check and illustrate let's add names for the vector items:
```{r}
names(v5) <- paste("No.", 1:10, sep="")
v5
```

Boolean values in R are called "logical":  
```{r}
class(v5)
is.logical(v5)
# is.logical(letters)
# typeof(letters)
# is.numeric(1:10)
# is.numeric(1.2)
```

## Data structures (2): Matrices  

Matrices are 2-dimensional arrays/tables where each item (think of Excel cells) has to be of the same data type:

```{r}
# generates a vector
1:24
```

Convert vector into a matrix with 6 columns  
```{r}
mx1 <- matrix(1:24, ncol=6)
mx1
```
```{r}
# Fill by column:
# mx2 <- matrix(LETTERS[1:24], nrow=6, byrow=FALSE)
# mx2
```

```{r}
# Fill by row:
# mx3 <- matrix(LETTERS[1:24], nrow=6, byrow=TRUE)
# mx3
```

## Calculations with vectors + matrices  

As we have already seen, one of the great things about vectors and matrices in R is that they can be modified with simple code:

```{r}
# v3
v3 * 100
```
```{r}
# mx1
mx1 + 1000
```
```{r}
# Transpose a matrix with 't()' function:
# mx1
# t(mx1)
```

## Data structures (3): Data frames  

Data frames are the most common container used to store data in R.  
Each column can be a different data type:  

```{r}
# 'iris' is another variable 'built into' R that stores a data frame:
# class(iris)
# class(mx1)
```

Generating a data frame 'by hand':
```{r}
df1 <- data.frame(Name=c("Jane", "Jack", "Jaim"),
                  Age=c(5, 12, 30),
                  Female=c(TRUE, FALSE, FALSE))
df1
```

To show what data type is found in each column of the data frame use this code:
```{r}
sapply(df1, class)
```

Very large data frames are best viewed in a separate tab:  
```{r}
View(df1)
```


### Subsetting (2): Subsetting data frames  

```{r}
print(df1)
```

Similar to vectors, extract data from data frames using square brackets;  
rows and columns are separated by a comma:  

* Rows: number before comma  
* Columns: number after comma  

Equivalent to "A1" (column A, row 1) in Excel:  
```{r}
df1[1,1]
```

Equivalent to "B3" (column B, row 3) in Excel:  
```{r}
df1[3,2]
```

Leaving the space before the comma blank will return entire columns (all rows):  
```{r}
df1[,1]
```

Accordingly, leaving the space AFTER the comma blank will return entire rows:
```{r}
df1[1,]
```


As with vectors, extract several rows/columns using number/index vectors:

* Before comma: `c(2,3)` = rows 2 and 3  
* After comma: `c(1,3)` = columns 1 and 3  

```{r}
df1[c(2,3), c(1,3)]
```
And again, as wwith vectors, we can also subset with Boolean/logical values:  

* Before comma: **`c(FALSE, TRUE, TRUE)`** =  
  - Row 1 = FALSE (don't include)  
  - Row 2 = TRUE (***do*** include)  
  - Row 3 = TRUE (***do*** include)  
* After comma: **`c(TRUE, FALSE, TRUE)`** =  
  - Column 1 = TRUE (***do*** include)  
  - Column 2 = FALSE (don't include)  
  - Column 3 = TRUE (***do*** include)  
  
```{r}
df1[c(FALSE, TRUE, TRUE), c(TRUE, FALSE, TRUE)]
```

### Column names/headers

For data frames the `names()` command refers to column headers:
```{r}
names(df1)
```


```{r}
names(df1) <- c("Nom", "Age", "Femelle")
df1
```

We often use the dollar sign to refer to entire column of a data frame (returns a vector):
```{r}
df1$Nom
class(df1$Nom)
```

Data frames (and matrices) can have row names as well:
```{r}
row.names(df1) <- paste0("Row", seq_len(nrow(df1)))
df1
```

And similar to vectors, we can use row names and column names to subset data frames:  
*(NOTE that text always needs to be defined with quotation marks)*  

Select 2 columns (all rows):
```{r}
df1[, c("Nom", "Femelle")]
```

Select rows (all columns):
```{r}
df1[c("Row1", "Row3"), ]
```

Select columns and rows:
```{r}
df1[c("Row1", "Row3"), c("Nom", "Femelle")]
```
Sometimes code becomes easier to read if index vectors are first stored in a variable; then use the variable to subset the data frame:
```{r}
a <- c(2,3)
b <- c(TRUE, FALSE, TRUE)
c <- c("Nom", "Age")
```

```{r}
df1[a, b]
```

```{r}
df1[b, c]
```
### Merging ("joining") data frames  

We often need to combine 2 data frames. This can be done with the `merge()` function.  

The data frames to be merged must have at least one column in common. In the example below, both the df2a and df2b data frames share the "Gene" column. The values in the shared column do not have to be in the same order.  
```{r}
df2a <- data.frame(
        Gene = c("ABCG1", "IRF1", "STAT3", "NEU1"),
        Gene.ID = c(9619, 3659, 6774, 4758)
)
df2a
```


```{r}
df2b <- data.frame(
        Gene = c("ABCG1", "IRF1", "STAT3", "NEU1"),
        Gene.Name = c("ATP binding cassette subfamily G member 1",
                      "interferon regulatory factor 1",
                      "signal transducer and activator of transcription 3",
                      "neuraminidase 1"
                      )
)
```


```{r}
df3 <- merge(x=df2a, y=df2b, by="Gene")
df3
```


